Penalized linear mixed models for structured genetic data

### 2021-03-17 main.tex
modifying figure 5
Edits to intro, software section, genome-wide section. replacing eps with pdfs

\pb{Read this over and let me know what you think}

Genome-wide association studies (GWAS) have become widespread over the past fifteen years, successfully identifying thousands of genetic variants associated with diseases and complex traits. Accurately identifying and estimating the effect sizes of these variants or single nucleotide polymorphisms (SNPs) is of interest for a variety of scientific applications including the prediction of an individual's risk of disease, understanding the genetic architecture underlying a complex trait, and drug discovery. Genetic association is typically analyzing by testing individual variants one at a time.  While straightforward, there are inherent limitations to analyzing each variant in isolation including diminished power, an inability to examine epistatic interactions, and bias due to the presence of correlation between variants.

Correlation among genetic loci, or linkage disequilibrium (LD), causes true biological associations to be mis-attributed to nearby, correlated SNPs. If one is analyzing each variant in isolation, then additional steps need to be taken to avoid this bias.  For example, in LD pruning, SNPs that exceed a correlation threshold are sequentially removed until only variants that are relatively independent are used in analysis and testing procedures. An alternative solution is the use of multivariate methods such as multiple linear regression, which adjusts for and estimates the effects of many variants simultaneously. However, classical regression relies on data with more many more observations than variants. This is typically infeasible in genome-wide studies where the variants to be analyzed number in the hundreds of thousands or even millions.

@@

Lasso models, however, rely on an assumption that subjects are independent of each other. In the presence of population structure or cryptic relatedness, however, this assumption is violated: traits are more likely to be similar within a family or ethnic group. The presence of structure can adversely affect analyses in several ways, but the most important concern is that it is likely to introduce false positives. Allele frequency differences due to ancestry within a sample are likely to be mistaken for biological signal if steps are not taken to mitigate this effect. One common solution, dividing a sample and performing ancestry-specific analyses, often results in a loss of statistical power from reduced sample sizes. Furthermore, ancestry-specific analyses cannot easily be applied in the presence of admixture.

@@

In order to clarify the subtle distinction between population structure and environmental heterogeneity and illustrate its importance, we present a detailed review of relevant concepts and methods. The remainder of this paper is organized as follows. In Section \ref{sec:background} we provide a summary of the causes and consequences of structure in the genetic data of seemingly unrelated individuals and formally define terminology used throughout this work. We emphasize that confounding due to population structure, frequently cited as a driver of spurious associations, is a function of environmental heterogeneity, where genetic data serves as a proxy for differential environmental exposures \citep{Sillanpaeae2011, sul2018population, vilhjalmsson2012nature, barton2019population}. In so doing, we refine the definition of confounding due to population structure in terms of a non-genetic mechanism. In addition, we briefly review historical methods of correcting for population stratification and relatedness. In Section \ref{sec:methods} we provide a detailed review of the two most prevalent multivariate methods of adjusting for population stratification and relatedness: PCA adjusted lasso penalized regression (PC-lasso) and lasso penalized multivariate LMMs (LMM-lasso). We review the statistical details of how these methods attempt to correct for population structure and environmental heterogeneity. In Section \ref{sec:results}, we illustrate these concepts via simulation studies and identify scenarios where particular methods are most and least effective in accurately estimating the effect sizes of observed SNPs in the presence of unobserved confounding. Finally, in Section~\ref{sec:gw_cs} we illustrate the use of these methods in a case study involving genome-wide data involving a sample of diverse racial background.

@@

\pb{Please read the following paragraph; it has been modified extensively.}

In many genetic association studies the phenotype of interest is binary, for example the presence or absence of a heritable trait. Unfortunately, the extension of LMMs and penalized LMMs to nonlinear models is not straightforward. As we explain in Section~\ref{sec:methods}, LMMs and penalized LMMs are able to efficiently correct for dependencies among subjects by preprocessing the data by means of a weighted rotation. This procedure does not directly generalize beyond the linear regression case. The situation is certainly not impossible, and many approaches to fitting generalized linear mixed models, or GLMMs, have been proposed in the classical, low-dimensional setting. However, very little work has been done to extend this to penalized, high-dimensional models and to the best of our knowledge there is no existing software for doing so at this time. We hope that this situation changes in the near future.

\pb{What is the significance of this Mandt paper? Isn't there a huge literature on GLMMs?}\anna{ \citet{mandt2017sparse} generalize the \textbf{high-dimensional, penalized} LMM modeling paradigm to binary phenotypes in the interest of classification. They do so by thresholding the outcome variable through a Probit likelihood with additional sparsity constraints and approximate Bayesian inference. }

@@

All analyses and simulations were performed using R version 3.6.3. The ordinary lasso and PC-lasso methods were implemented using \texttt{glmnet} \citep{glmnet}. As described previously, LMM-lasso-ggmix requires specialized software, but is implemented in the R package \texttt{ggmix} \citep{bhatnagar2020simultaneous}. Standard lasso software such as \texttt{glmnet} can be used to fit the LMM-lasso-Rakitsch method as well, although this requires a fair amount of preprocessing and setting up. To make this process easier, we developed an R package \texttt{penalizedLMM}, available at the github repository \url{https://github.com/areisett/penalizedLMM}, to allow seamless, user-friendly fitting of penalized LMMs based on genotype matrices or PLINK files. \anna{\texttt{penalizedLMM} includes preprocessing functions for \anna{SNP and sample-level filtering and} genotype imputation using \textbf{imputation rules - how to describe this} where possible, or else the Hardy-Weinberg Equilibrium value.} \texttt{penalizedLMM} additionally implements the elastic net, and the nonconvex penalties, MCP and SCAD. \anna{refine filtering, imputation functions}

@@

\begin{figure}[H]
     \centering
     \begin{subfigure}[b]{0.475\textwidth}
         \centering
         \includegraphics[width=\textwidth]{figures/figure_05.png}
         \caption{Precision for each method and number of selected SNPs from the Empirical genome-wide analysis. Single replication case study with $\eta = 0.5, \xi = 0.8$.}
        \label{fig:gw_prc}
     \end{subfigure}
     \hfill
     \begin{subfigure}[b]{0.475\textwidth}
         \centering
         \includegraphics[width=\textwidth]{figures/figure_06.png}
         \caption{MSE for each method and number of selected SNPs from the Empirical genome-wide analysis. Single replication case study with $\eta = 0.5, \xi = 0.8$.}
        \label{fig:gw_mse}
     \end{subfigure}
\end{figure}

>>>>

\anna{Overall I think this looks good and definitely more friendly. Fixed some minor typos, repetitive words (e.g. 'however' twice in a sentence)}

Genome-wide association studies (GWAS) have become widespread over the past fifteen years, successfully identifying thousands of genetic variants associated with diseases and complex traits. Accurately identifying and estimating the effect sizes of these variants or single nucleotide polymorphisms (SNPs) is of interest for a variety of scientific applications including the prediction of an individual's risk of disease, understanding the genetic architecture underlying a complex trait, and drug discovery. Genetic association is typically assessed by testing individual variants one at a time.  While straightforward, there are inherent limitations to analyzing each variant in isolation including diminished power, an inability to examine epistatic interactions, and bias due to the presence of correlation between variants.

Correlation among genetic loci, or linkage disequilibrium (LD), causes true biological associations to be mis-attributed to nearby, correlated SNPs. If one is analyzing each variant in isolation, then additional steps need to be taken to avoid this bias.  For example, in LD pruning, SNPs that exceed a correlation threshold are sequentially removed until only variants that are relatively independent are used in analysis and testing procedures. An alternative solution is the use of multivariate methods such as multiple linear regression, which adjusts for and estimates the effects of many variants simultaneously. However, classical regression relies on data with many more observations than variants. This is typically infeasible in genome-wide studies where the variants to be analyzed number in the hundreds of thousands or even millions.

@@

Lasso models, however, rely on an assumption that subjects are independent of each other. In the presence of population structure or cryptic relatedness this assumption is violated: traits are more likely to be similar within a family or ethnic group. The presence of structure can adversely affect analyses in several ways, but the most important concern is that it is likely to introduce false positives. Allele frequency differences due to ancestry within a sample are likely to be mistaken for biological signal if steps are not taken to mitigate this effect. One common solution, dividing a sample and performing ancestry-specific analyses, often results in a loss of statistical power from reduced sample sizes. Furthermore, ancestry-specific analyses cannot easily be applied in the presence of admixture.

@@

In order to clarify the subtle distinction between population structure and environmental heterogeneity and illustrate its importance, we present a detailed review of relevant concepts and methods. The remainder of this paper is organized as follows. In Section \ref{sec:background} we provide a summary of the causes and consequences of structure in the genetic data of seemingly unrelated individuals and formally define terminology used throughout this work. We emphasize that confounding due to population structure, frequently cited as a driver of spurious associations, is a function of environmental heterogeneity, where genetic data serves as a proxy for differential environmental exposures \citep{Sillanpaeae2011, sul2018population, vilhjalmsson2012nature, barton2019population}. In so doing, we refine the definition of confounding due to population structure in terms of a non-genetic mechanism. In addition, we briefly review historical methods of correcting for population stratification and relatedness. In Section \ref{sec:methods} we provide a detailed review of the two most prevalent multivariate methods of adjusting for population stratification and relatedness: PCA adjusted lasso penalized regression (PC-lasso) and lasso penalized multivariate LMMs (LMM-lasso). We review the statistical details of how these methods attempt to correct for population structure and environmental heterogeneity. In Section \ref{sec:results}, we illustrate these concepts via simulation studies and identify scenarios where particular methods are most and least effective in accurately estimating the effect sizes of observed SNPs in the presence of unobserved confounding. Finally, in Section~\ref{sec:gw_cs} we illustrate the use of these methods in a case study of genome-wide data involving a sample of diverse racial background.

@@

In many genetic association studies the phenotype of interest is binary, for example the presence or absence of a heritable trait. Unfortunately, the extension of LMMs and penalized LMMs to nonlinear models is not straightforward. As we explain in Section~\ref{sec:methods}, LMMs and penalized LMMs are able to efficiently correct for dependencies among subjects by preprocessing the data by means of a weighted rotation. This procedure does not directly generalize beyond the linear regression case. The situation is certainly not impossible, and many approaches to fitting generalized linear mixed models, or GLMMs, have been proposed in the classical, low-dimensional setting. \anna{However, very little work has been done to extend this to penalized, high-dimensional models. As far we are aware, only \citet{mandt2017sparse} generalize the high-dimensional penalized LMM modeling paradigm to binary phenotypes. They do so by thresholding the outcome variable through a Probit likelihood with additional sparsity constraints and approximate Bayesian inference. However, to the best of our knowledge there is no available software to implement this or any other methods for high-dimensional, penalized methods at this time.} We hope that this situation changes in the near future.

\pb{What is the significance of this Mandt paper? Isn't there a huge literature on GLMMs?}\anna{Mandt is specifically for high-dimensional, penalized LMMs that use the data to infer structure of the errors - not just a regular GLMM. Includes some of the Rakitsch 2013 authors, extends LMM-lasso to binary outcomes in one way.}

@@

All analyses and simulations were performed using R version 3.6.3. The ordinary lasso and PC-lasso methods were implemented using \texttt{glmnet} \citep{glmnet}. As described previously, LMM-lasso-ggmix requires specialized software, but is implemented in the R package \texttt{ggmix} \citep{bhatnagar2020simultaneous}. Standard lasso software such as \texttt{glmnet} can be used to fit the LMM-lasso-Rakitsch method as well, although this requires a fair amount of preprocessing and setting up. To make this process easier, we developed an R package \texttt{penalizedLMM}, available at the github repository \url{https://github.com/areisett/penalizedLMM}, to allow seamless, user-friendly fitting of penalized LMMs based on genotype matrices or PLINK files. \anna{\texttt{penalizedLMM} includes preprocessing functions for genotype imputation using \textbf{imputation rules - how to describe this} where possible, or else the Hardy-Weinberg Equilibrium value.} \texttt{penalizedLMM} additionally implements the elastic net, and the nonconvex penalties, MCP and SCAD.

@@

\begin{figure}[H]
     \centering
     \begin{subfigure}[b]{0.4\textwidth}
         \centering
         \includegraphics[width=\textwidth]{figures/figure_05a.pdf}
         \caption{Precision for each method and number of selected SNPs from the Empirical genome-wide analysis. Single replication case study with $\eta = 0.5, \xi = 0.8$.}
        \label{fig:gw_prc}
     \end{subfigure}
    \hfill
     \begin{subfigure}[b]{0.4\textwidth}
         \centering
         \includegraphics[width=\textwidth]{figures/figure_05b.pdf}
         \caption{MSE for each method and number of selected SNPs from the Empirical genome-wide analysis. Single replication case study with $\eta = 0.5, \xi = 0.8$.}
        \label{fig:gw_mse}
     \end{subfigure}
     \begin{subfigure}[b]{0.15\textwidth}
         \centering
         \includegraphics[width=\textwidth]{figures/figure_05c.pdf}
     \end{subfigure}
\end{figure}
